---
title: "Project"
author: "Sierra Hess"
date: "11/9/2022"
output: html_document
---
Set things up
```{r}
rm(list = ls())
library(tidyverse)
library(here)
library(ggfortify)
```
Now let's pull in the data:
```{r}
dry_mass <- read.csv(here("Data", "dry_mass_metadata.csv"),stringsAsFactors = TRUE)
germination <- read.csv(here("Data", "germination_data.csv"),stringsAsFactors = TRUE)
```
# Before we begin any analysis, we need to check our data for possible errors/outliers.

Species Names:
```{r}
levels(dry_mass$species)
levels(germination$species)
```
This looks good, so now we can look at relationships.

## We can start off by looking at histograms for the data. 
```{r}
ggplot(dry_mass, aes(x = mass))+
     geom_histogram()
```

This plot looks reasonable with no obvious outliers.

## Now check soil moisture
```{r}
ggplot(germination, aes(x = soil_moisture))+
     geom_histogram()
```

This plot also looks reasonable. 

## Let's also just run a summary on all the variables:
```{r}
summary(dry_mass)
summary(germination)
```

From the summary table, there are no clear issues with the dataset. 

# Now we can look at the soil moisture relation to each species
Because we have count data, using a bar graph is okay.

```{r}
ggplot(germination, aes(x = species, y = number_germinated))+
     geom_bar(stat= "identity", position = "dodge")
```

We can see that the Salvia columbarie species had low germination rates while the other three species had much a much higher number of germinated plants.

Species is a categorical variable and the number of plants germinated is a discrete variable, so the likely test is a Chi-square test for independence. In order to run this test, we need our data summarized in a matrix.

```{r}
germ.mat <- xtabs(number_germinated ~ species, data = germination)
germ.mat
```
Now we can run the test
```{r}
chisq.test(germ.mat)
```

Our null hypothesis is that there is no difference in the number of germinated plants between species. However,  we can reject our null hypothesis given that our p value is less than .05, so there is a statistically significant difference between the number of germinated plants between species.

# Next we can look at how mass varies between species

Let's first plot the data by using a boxplot.

```{r}
ggplot(dry_mass, aes(species, mass))+
     geom_boxplot()+
     geom_point(alpha = 0.5)
```

There is a point that appears fairly distant from the mean in the Phacelia tanacetifolia plot, but there isn't a clear outlier.


Now we can put the data into matrix form so that we can run a chisq.test
```{r}
mass.mat <- xtabs(mass ~ species, data = dry_mass)
mass.mat
```
Now we can run the test
```{r}
chisq.test(mass.mat)
```

Since we have a low p value, we can reject the null hypothesis and there is a clear relationship between the species and mass. 

# Lets look at levels and mass
```{r}
low_med_data<-dry_mass %>% filter(factor == "Low" | factor == "Medium")

low_med_data$factor <- factor(low_med_data$factor)
```

Lets look at a figure of the data

```{r}
ggplot(low_med_data, aes(factor, mass))+
     geom_boxplot()+
     geom_point(position = "jitter", size = 2, alpha = 0.5)+
     theme_bw()
```

The plots look similar, but lets do a formal test. 

Now let's do a t-test
```{r}
MyT<-t.test(mass ~ factor, data = low_med_data)
MyT
```
We cannot reject the null hypothesis given that p > .05, so there is no clear difference between mass for low and medium factors.

## Now for medium and high

```{r}
med_high_data<-dry_mass %>% filter(factor == "Medium" | factor == "High")

med_high_data$factor <- factor(med_high_data$factor)
```

Lets look at a figure of the data

```{r}
ggplot(med_high_data, aes(factor, mass))+
     geom_boxplot()+
     geom_point(position = "jitter", size = 2, alpha = 0.5)+
     theme_bw()
```

The plots look to be slightly different, but lets do a formal test. 

Now let's do a t-test
```{r}
MyT2<-t.test(mass ~ factor, data = med_high_data)
MyT2
```
We cannot reject the null hypothesis given that p > .05, so there is no clear difference between mass for medium and high factors.

## Now lets do low vs high factors

```{r}
low_high_data<-dry_mass %>% filter(factor == "Low" | factor == "High")

low_high_data$factor <- factor(low_high_data$factor)
```

Lets look at a figure of the data

```{r}
ggplot(low_high_data, aes(factor, mass))+
     geom_boxplot()+
     geom_point(position = "jitter", size = 2, alpha = 0.5)+
     theme_bw()
```

The plots look similar, but lets do a formal test. 

Now let's do a t-test
```{r}
MyT<-t.test(mass ~ factor, data = low_high_data)
MyT
```
We cannot reject the null hypothesis given that p > .05, so there is no clear difference between mass for low and high factors. However, the p value was lowest for low vs high, so there is the highest level of difference between low and high factors when compared to high and med and low and med. 

